Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-10-20 TACO-paper-v2/coffee.tex
edits

In order to overcome the constraints of the available compilers and
specialized linear algebra libraries, we have automated a set of generic
and model-driven code transformations in COFFEE\footnote{COFFEE stands
for COmpiler For FinitE Element local assembly.}, a compiler for
optimizing local assembly kernels. COFFEE is integrated with
Firedrake~\cite{firedrake-code}, a system for solving PDEs through the
finite element method based on the PyOP2
abstraction~\cite{pyop2ws,pyop2isc}. All problems expressible with this
framework are supported by COFFEE, including equations that can be found
at the core of real-world simulations, such as those used in our
performance evaluation. In evaluating our code transformations for a
range of relevant problems, we vary two key parameters that impact
solution accuracy and kernel cost: the polynomial order of the method
(we investigate from $p=1$ to $p=4$) and the geometry of elements in the
discretized domain (2D triangle, 3D tetrahedron, 3D prism). From the
point of view of the generated code, these two parameters directly
impact the size of both loop nests and mathematical expressions, as
elaborated in Section~\ref{sec:real-examples}.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz}, which shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
unique arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionally to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

@@

Splitting an expression (henceforth \emph{split}) has, however, several drawbacks. Firstly, it
increases the number of accesses to $A$ proportionally to the ``split
factor,'' which is the number of sub-expressions produced. Also,
depending on how the split is executed, it can lead to redundant
computation. For example, the number of times the product $det*W3[i]$ is
performed is proportional to the number of sub-expressions, as shown in
the code snippet. Further, it increases loop overhead, for example
through additional branch instructions. Finally, it might affect
register locality: for instance, the same array could be accessed in
different sub-expressions, requiring a proportional number of loads be
performed. This is not the case for the Helmholtz example. Nevertheless,
as shown in Section~\ref{sec:perf-results}, the performance gain from
improved register reuse along inner dimensions can still be greater,
especially if the split factor and the splitting itself use heuristics
to minimize the aforementioned issues.
~\\

Some of the transformations presented in this section, as well as other typical compiler optimizations applicable by COFFEE, like loop interchange and unrolling, are parametric; for instance, expression splitting and loop unrolling depend on the split and the unroll factors, respectively. Since all transformations are composable, the optimization space can be as large as order of thousands variants for local assembly kernels in which mathematical expressions are moderately complex and the iteration space is not particularly small, so a strategy to prune it is required. This problem is treated in the following section.

@@

COFFEE uses heuristics and a simple cost model to select the most suitable
optimization strategy for a given problem. Auto-tuning might be used, although
it would significantly increase the  run-time overhead, since the generation of
ASTs occurs at run-time as soon as problem-specific data are available.

@@

Experiments were run on a single core of two Intel architectures, a
Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB
L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode,
32KB L1 cache and 512KB L2 cache). We have chosen these two
architectures because of the differences in the number of logical
registers and SIMD lanes (16 256-bit registers in the Sandy Bridge,
and 32 512-bit registers in the Xeon Phi), which can impact the
optimization strategy. The \texttt{icc 13.1} compiler was used. 
We selected the best optimization levels for each platform using the
following compilation flags: \texttt{-xAVX} for auto-vectorization and
\texttt{-O2} were used on the Sandy Bridge; \texttt{-O3} was adopted
on the Xeon Phi.

@@

The gain tends to grow with the computational cost of the kernels:
bigger loop nests (i.e., larger element shapes and polynomial orders)
usually benefit from the reduction in redundant computation, even
though extra memory for the temporary arrays is required. Some
discrepancies to this trend are due to a less effective
auto-vectorization. For instance, on the Sandy Bridge, the improvement
at $p=3$ is larger than that at $p=4$ because, in the latter case, the
size of the innermost loop is not a multiple of the vector length, and
a remainder scalar loop is introduced at compile time. Since the loop
nest is small, the cost of executing the extra scalar iterations can
have a significant impact. The remainder loop overhead is more
pronounced on the Xeon Phi, where the vector length is twice as long,
which leads to proportionally larger scalar remainder loops.

@@

In these experiments, we observe a maximum performance improvement of
1.47$\times$ over the non-optimized local assembly code, obtained in the
case ($f=2, p=2$). However, we reiterate the fact that full application
speed ups increase proportionally with the amount of time spent in
assembly and, therefore, with the complexity of the equation. By
increasing polynomial order and number of coefficient functions, or by
simply studying a different, more complex equation, it is our experience
that performance gains become increasingly more relevant. The choice of
studying the static linear elasticity equation was to show that even
relatively simple problems can be characterized by a large proportion of
execution time spent in assembly.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for improving
arithmetic intensity in finite element local assembly kernels, and their
integration in a novel compiler, COFFEE. In the context of automated code
generation for finite element local assembly, COFFEE is the first compiler
capable of introducing low-level optimizations to maximize instruction-level
parallelism, register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very small, with
the size depending on aspects like the degree of accuracy one wants to
reach (polynomial order of the method) and the mesh discretization employed.
The data space, in terms of number of arrays and scalars required to evaluate
the element matrix, grows proportionally with the complexity of the
finite element problem. COFFEE has been developed taking into account
all of these degrees of freedom, based on the idea that reducing the
problem of local assembly optimization to a fixed sequence of
transformations is far too superficial if close-to-peak performance
needs to be reached. The various optimizations overcome limitations of
current vendor and research compilers. The exploitation of domain
knowledge allows some of them to be particularly effective, as
demonstrated by our experiments on two state-of-the-art Intel
platforms. Further work includes a comprehensive study about
feasibility and constraints on transforming kernels into a sequence of
calls to external linear algebra libraries. COFFEE supports all of the
problems expressible in Firedrake, and is already integrated with
this framework. The generality and the applicability of the proposed
code transformations to other domains has also been discussed.

>>>>

In order to overcome the constraints of the available compilers and
specialized linear algebra libraries, we have automated a set of generic
and model-driven code transformations in COFFEE\footnote{COFFEE stands
for COmpiler For Fast Expression Evaluation.}, a compiler for
optimizing local assembly kernels. COFFEE is integrated with
Firedrake~\cite{firedrake-code}, a system for solving PDEs through the
finite element method based on the PyOP2
abstraction~\cite{pyop2ws,pyop2isc}. All problems expressible with this
framework are supported by COFFEE, including equations that can be found
at the core of real-world simulations, such as those used in our
performance evaluation. In evaluating our code transformations for a
range of relevant problems, we vary two key parameters that impact
solution accuracy and kernel cost: the polynomial order of the method
(we investigate from $p=1$ to $p=4$) and the geometry of elements in the
discretized domain (2D triangle, 3D tetrahedron, 3D prism). From the
point of view of the generated code, these two parameters directly
impact the size of both loop nests and mathematical expressions, as
elaborated in Section~\ref{sec:real-examples}.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz}, which shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
unique arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionately to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

@@

Splitting an expression (henceforth \emph{split}) has, however,
several drawbacks. Firstly, it increases the number of accesses to $A$
in proportion to the ``split factor,'' which is the number of
sub-expressions produced. Also, depending on how splitting is done, it
can lead to redundant computation. For example, the number of times
the product $det*W3[i]$ is performed is proportional to the number of
sub-expressions, as shown in the code snippet. Further, it increases
loop overhead, for example through additional branch
instructions. Finally, it might affect register locality: for
instance, the same array could be accessed in different
sub-expressions, requiring a proportional number of loads be
performed. This is not the case for the Helmholtz
example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the
performance gain from improved register reuse along inner dimensions
can still be greater, especially if the split factor and the splitting
itself use heuristics to minimize the aforementioned issues.

@@

Some of the transformations presented in the previous section, as well
as other typical compiler optimizations used in COFFEE, like loop
interchange and unrolling, are parametric; for instance, expression
splitting and loop unrolling depend on the split and the unroll
factors, respectively. Since all transformations are composable, the
optimization space can be as large as order of thousands variants for
local assembly kernels in which mathematical expressions are
moderately complex and the iteration space is not particularly small,
so a strategy to prune it is required.  For this, COFFEE uses
heuristics and a simple cost model to select the most suitable
optimization strategy for a given problem. Auto-tuning can also be
used, although it would significantly increase the run-time overhead,
since the generation of ASTs occurs at run-time as soon as
problem-specific data are available.

@@

Experiments were run on a single core of two Intel architectures, a
Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB
L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode,
32KB L1 cache and 512KB L2 cache). We have chosen these two
architectures because of the differences in the number of logical
registers and SIMD lanes (16 256-bit registers in the Sandy Bridge,
and 32 512-bit registers in the Xeon Phi), which can impact the
optimization strategy. The \texttt{icc 13.1} compiler was used.  We
selected the best optimization levels for each platform; we used
\texttt{-xAVX} for auto-vectorization and \texttt{-O2} on the Sandy
Bridge, and \texttt{-O3} on the Xeon Phi.

@@

The gain tends to grow with the computational cost of the kernels:
bigger loop nests (i.e., larger element shapes and polynomial orders)
usually benefit from the reduction in redundant computation, even
though extra memory for the temporary arrays is required. Some
discrepancies to this trend are due to a less effective
auto-vectorization. For instance, on the Sandy Bridge, the improvement
at $p=3$ is larger than that at $p=4$ because, in the latter case, the
size of the innermost loop is not a multiple of the vector length, and
a remainder scalar loop is introduced at compile time. Since the loop
nest is small, the cost of executing the extra scalar iterations can
have a significant impact. The remainder loop overhead is more
pronounced on the Xeon Phi, where the vector length is twice as long,
which leads to proportionately larger scalar remainder loops.

@@

In these experiments, we observe a maximum performance improvement of
1.47$\times$ over the non-optimized local assembly code, obtained in the
case ($f=2, p=2$). However, we reiterate the fact that full application
speed ups increase proportionately with the amount of time spent in
assembly and, therefore, with the complexity of the equation. By
increasing polynomial order and number of coefficient functions, or by
simply studying a different, more complex equation, it is our experience
that performance gains become increasingly more relevant. The choice of
studying the static linear elasticity equation was to show that even
relatively simple problems can be characterized by a large proportion of
execution time spent in assembly.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for improving
arithmetic intensity in finite element local assembly kernels, and their
integration in a novel compiler, COFFEE. In the context of automated code
generation for finite element local assembly, COFFEE is the first compiler
capable of introducing low-level optimizations to maximize instruction-level
parallelism, register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very small, with
the size depending on aspects like the degree of accuracy one wants to
reach (polynomial order of the method) and the mesh discretization employed.
The data space, in terms of number of arrays and scalars required to evaluate
the element matrix, grows proportionately with the complexity of the
finite element problem. COFFEE has been developed taking into account
all of these degrees of freedom, based on the idea that reducing the
problem of local assembly optimization to a fixed sequence of
transformations is far too superficial if close-to-peak performance
needs to be reached. The various optimizations overcome limitations of
current vendor and research compilers. The exploitation of domain
knowledge allows some of them to be particularly effective, as
demonstrated by our experiments on two state-of-the-art Intel
platforms. Further work includes a comprehensive study about
feasibility and constraints on transforming kernels into a sequence of
calls to external linear algebra libraries. COFFEE supports all of the
problems expressible in Firedrake, and is already integrated with
this framework. The generality and the applicability of the proposed
code transformations to other domains has also been discussed.
